---
title: "P8106: Severe Flu Prediction Project"
author:
  - "Naomi Simon-Kumar"
  - ns3782
date: "05/10/2025"
output:
  pdf_document:
    latex_engine: xelatex
    toc: true
    toc_depth: 2
fontsize: 11pt
linestretch: 1
---


```{r setup, include=FALSE}

# Ensures R code is suppressed
knitr::opts_chunk$set(
  echo = FALSE,         
  warning = FALSE,      
  message = FALSE,      
  results = 'hide'      
)

```

```{r}

# Load libraries
library(tidyverse)
library(caret)
library(ggplot2)  
library(patchwork)
library(corrplot)
library(table1)
library(pROC)
library(MASS)
library(factoextra)
library(gridExtra)
library(corrplot)
library(RColorBrewer)
library(gplots)
library(jpeg)
library(visdat)
library(vip)
library(tidymodels)
library(knitr)

# Set seed for reproducibility
set.seed(299)

# Load dataset
flu_df <- read_csv("severe_flu.csv")

# Convert categorical variables to factors
flu_df <- flu_df %>%
  mutate(
    gender = factor(gender, levels = c(0, 1), labels = c("Female", "Male")),
    race = factor(race, levels = c(1, 2, 3, 4), labels = c("White", "Asian", "Black", "Hispanic")),
    smoking = factor(smoking, levels = c(0, 1, 2), labels = c("Never", "Former", "Current")),
    diabetes = factor(diabetes, levels = c(0, 1), labels = c("No", "Yes")),
    hypertension = factor(hypertension, levels = c(0, 1), labels = c("No", "Yes")),
    severe_flu = factor(severe_flu, levels = c(0, 1), labels = c("No", "Yes"))
  )

# Drop ID
flu_df <- dplyr::select(flu_df, -id)

# Set 10-fold cross-validation
ctrl1 <- trainControl(method = "cv", number = 10)

# Split data into testing and training
data_split <- initial_split(flu_df, prop = 0.8)
training_data <- training(data_split)
testing_data <- testing(data_split)

```

# 1. Introduction

## 1.1. Background and Study Objectives

This study aims to build prediction models for identifying factors associated with the incidence of severe flu within 6 months post-vaccination. The analysis focuses on a dataset of 1,000 participants to understand risk factors for severe flu in a population administered the flu vaccine.

The primary objectives of this project are to: (1) evaluate whether advanced predictive modeling techniques like boosting and support vector machines (SVM) provide superior predictive performance compared to simpler models, (2) develop a predictive risk score that quantifies the probability of experiencing severe flu based on individual characteristics, and (3) identify key demographic and clinical factors that predict severe flu risk and assess how these factors influence this risk.

## 1.2. Data Source and Description

The `severe_flu` dataset used for analysis contains demographic and clinical information from 1,000 participants. Variables include age (in years), gender (1 = Male, 0 = Female), race/ethnicity (1 = White, 2 = Asian, 3 = Black, 4 = Hispanic), smoking status (0 = Never smoked, 1 = Former smoker, 2 = Current smoker), height (in centimeters), weight (in kilograms), BMI (Body Mass Index, calculated as weight in kg divided by height in meters squared), diabetes status (0 = No, 1 = Yes), hypertension status (0 = No, 1 = Yes), systolic blood pressure (SBP, in mmHg), and LDL cholesterol (in mg/dL). The outcome of interest is a binary variable severe_flu indicating whether a participant experienced severe flu within 6 months post-vaccination (0 = No, 1 = Yes).

# 2. Exploratory Analysis

Exploratory data analysis was undertaken to examine the structure of the `severe_flu` dataset, including assessment of data distributions, and relationships between variables. The dataset contains 1,000 observations with 12 variables after removing the ID column. Categorical variables (gender, race, smoking status, diabetes, hypertension, and severe flu outcome) were converted to factors with appropriate labels. The dataset was verified to be complete with no missing observations.

Summary statistics were calculated for all demographic and clinical variables. Visualizations were produced to compare characteristics between participants who did and did not experience severe flu.

## 2.1. Summary Statistics

**Table 1. Summary statistics of demographic and clinical characteristics in the study population ** 
```{r results='asis'}

# Set seed for reproducibility
set.seed(299)

# Reference code: https://cran.r-project.org/web/packages/table1/vignettes/table1-examples.html 

label(flu_df$age) <- "Age (yrs)"
label(flu_df$gender) <- "Gender"
label(flu_df$race) <- "Race/Ethnicity"
label(flu_df$smoking) <- "Smoking Status"
label(flu_df$height) <- "Height (cm)"
label(flu_df$weight) <- "Weight (kg)"
label(flu_df$bmi) <- "Body Mass Index (weight/ (height)^2)"
label(flu_df$diabetes) <- "Diabetes"
label(flu_df$hypertension) <- "Hypertension"
label(flu_df$SBP) <- "Systolic Blood Pressure (mmHg)"
label(flu_df$LDL) <- "LDL cholesterol (mg/dL)"

table1= table1(~ gender + race + smoking + height + weight + bmi + diabetes + hypertension + SBP + LDL, data=flu_df)
knitr::kable(table1)

```

Table 1 summarises the demographic and clinical characteristics of the study population (N = 1000). The sample was balanced by gender (52.2% female, 47.8% male), with predominantly White participants (65.6%). Most participants were never smokers (58.4%), with 14.5% reporting diabetes and 46.4% identified as hypertensive. Notably, the mean BMI was 27.9 kg/m² (SD=2.76), indicating the average subject would be considered clinically overweight. Mean systolic blood pressure was 130 mmHg (SD=7.88), and mean LDL cholesterol was 110 mg/dL (SD=19.7).

## 2.2. Exploratory Plots

### 2.2.1. Histograms

```{r}

## Histograms for numeric predictors

h1 <- ggplot(training_data, aes(x = age)) +
  geom_histogram(binwidth = 2, color = "darkblue", fill = "lightblue") +
  ggtitle("Age Distribution") +
  ylab("") +
  xlab("") +
  theme(plot.title = element_text(hjust = 0.5))

h2 <- ggplot(training_data, aes(x = bmi)) +
  geom_histogram(binwidth = 1, color = "darkblue", fill = "lightblue") +
  ggtitle("BMI") +
  ylab("") +
  xlab("") +
  theme(plot.title = element_text(hjust = 0.5))

h3 <- ggplot(training_data, aes(x = SBP)) +
  geom_histogram(binwidth = 5, color = "darkblue", fill = "lightblue") +
  ggtitle("Systolic BP") +
  ylab("") +
  xlab("") +
  theme(plot.title = element_text(hjust = 0.5))

h4 <- ggplot(training_data, aes(x = LDL)) +
  geom_histogram(binwidth = 5, color = "darkblue", fill = "lightblue") +
  ggtitle("LDL Cholesterol") +
  ylab("") +
  xlab("") +
  theme(plot.title = element_text(hjust = 0.5))

h5 <- ggplot(training_data, aes(x = height)) +
  geom_histogram(binwidth = 2, color = "darkblue", fill = "lightblue") +
  ggtitle("Height") +
  ylab("") +
  xlab("") +
  theme(plot.title = element_text(hjust = 0.5))

h6 <- ggplot(training_data, aes(x = weight)) +
  geom_histogram(binwidth = 2, color = "darkblue", fill = "lightblue") +
  ggtitle("Weight") +
  ylab("") +
  xlab("") +
  theme(plot.title = element_text(hjust = 0.5))

# Combine plots
# Ref code: https://tidytales.ca/snippets/2022-12-22_patchwork-shared-axis-labels/#shared-y-axis-labels
combined_histogram_numeric <- (h1 + h2) / (h3 + h4) / (h5 + h6)
wrap_elements(combined_histogram_numeric) +
  labs(tag = "Count") +
  theme(
    plot.tag = element_text(size = rel(1), angle = 90),
    plot.tag.position = "left"
  )



```

**Figure 1. Distribution of continuous predictors in the study dataset**

Figure 1 presents histogram plots for numeric predictors in the study dataset. Age, BMI, SBP, LDL, height and weight are approximately normally distributed, with minimal right skew for LDL.


### 2.2.2. Boxplots


```{r}

# Boxplots of numeric predictors
# Ref code: https://patchwork.data-imaginist.com/reference/plot_annotation.html
theme_no_xlab <- theme_bw() + theme(axis.title.x = element_blank())
bp1 <- ggplot(training_data, aes(x = severe_flu, y = age)) + geom_boxplot() + theme_no_xlab
bp2 <- ggplot(training_data, aes(x = severe_flu, y = bmi)) + geom_boxplot() + theme_no_xlab
bp3 <- ggplot(training_data, aes(x = severe_flu, y = SBP)) + geom_boxplot() + theme_no_xlab
bp4 <- ggplot(training_data, aes(x = severe_flu, y = LDL)) + geom_boxplot() + theme_no_xlab
bp5 <- ggplot(training_data, aes(x = severe_flu, y = height)) + geom_boxplot() + theme_no_xlab
bp6 <- ggplot(training_data, aes(x = severe_flu, y = weight)) + geom_boxplot() + theme_no_xlab

# Combine all plots
boxplot_numeric <- (bp1 + bp2) / (bp3 + bp4) / (bp5 + bp6) 

# Ref code: https://tidytales.ca/snippets/2022-12-22_patchwork-shared-axis-labels/#shared-x-axis-labels 
wrap_elements(boxplot_numeric) +
  labs(tag = "Severe Flu") +
  theme(
    plot.tag = element_text(size = rel(1)),
    plot.tag.position = "bottom"
  )

```

**Figure 2. Distribution of continuous predictors by severe flu status in the study dataset** 

From the boxplots of continuous predictors, age, SBP, LDL, and height distributions appear similar between severe flu and non-severe flu groups, with considerable middle quartile overlap (Figure 2). BMI and weight show slightly higher median values in the severe flu group, however, the substantial overlap in their distributions suggest these differences may not be statistically significant.

### 2.2.2. Barplots

```{r}

# Bar Plots of Categorical Predictors by Outcome

bar1 <- ggplot(training_data, aes(x = gender, fill = severe_flu)) +
  geom_bar(position = "fill") + labs(y = NULL)  + labs(fill = "Severe Flu") + theme_bw()

bar2 <- ggplot(training_data, aes(x = race, fill = severe_flu)) +
  geom_bar(position = "fill") + labs(y = NULL)  + theme_bw() + theme(legend.position = "none")

bar3 <- ggplot(training_data, aes(x = smoking, fill = severe_flu)) +
  geom_bar(position = "fill") + labs(y = NULL)  + theme_bw() + theme(legend.position = "none")

bar4 <- ggplot(training_data, aes(x = diabetes, fill = severe_flu)) +
  geom_bar(position = "fill") + labs(y = NULL)  + theme_bw() + theme(legend.position = "none")

bar5 <- ggplot(training_data, aes(x = hypertension, fill = severe_flu)) +
  geom_bar(position = "fill") + labs(y = NULL) + theme_bw() + theme(legend.position = "none")

# Combine using patchwork
((bar1 + bar2) / (bar3 + bar4) / (bar5 + plot_spacer())) +
  plot_layout(guides = "collect")

```

**Figure 3. Proportion of severe flu outcomes across categorical variables in the study dataset** 

Among categorical predictors, there appears to be differences in severe flu rates (Figure 3). Hispanic subjects show a higher proportion of severe flu cases compared to other racial groups. Diabetic patients also appear to have a slightly higher severe flu rate than non-diabetic individuals. Similarly current smokers have a slightly higher proportion of severe flu cases compared to never and former smokers. Gender and hypertension status show minimal differences in severe flu proportions across categories.

## 2.3. Correlation Analysis


```{r}

# Set seed for reproducibility
set.seed(299)

## Correlation Plot

# Correlation plot for numeric variables
numeric_vars <- training_data %>% dplyr::select(age, height, weight, bmi, SBP, LDL)
corrplot(cor(numeric_vars), method = "circle") 

```

**Figure 4. Correlation matrix between continuous predictors in the study dataset**

The correlation plot shows several relationships among continuous predictors. Weight and BMI have a strong positive correlation ($r \approx 0.8$), which may indicate collinearity. This is expected as BMI is calculated directly from weight and height. Therefore, interpretation of model results, including identifying important predictors of severe flu, should consider that these variables overlap in what they measure. Age and SBP have a moderate positive correlation ($r \approx 0.4$). Height and weight are weakly positively correlated ($r \approx 0.2$). By comparison, height has a moderate negative correlation with BMI ($r \approx -0.4$).

# 3. Model Training

## 3.1. Data preprocessing

The data was examined to identify missing values and undertake data cleaning. No missing observations were identified. Categorical variables were converted into appropriate factor variables: gender (Female/Male), race/ethnicity (White, Asian, Black, Hispanic), smoking status (Never, Former, Current), diabetes (No/Yes), hypertension (No/Yes), and the response variable severe flu (No/Yes). The ID variable was removed as it was not a meaningful predictor for flu severity. The preprocessed dataset was then split into training (80%) and testing (20%) sets for model development and evaluation.

## 3.2. Model Evaluation metrics

Area Under the Receiver Operating Characteristic curve (AUC) was selected as the metric for evaluating classification performance across all candidate models. AUC measures a model's ability to discriminate between severe and non-severe flu cases across all possible classification thresholds, with values ranging from 0.5 (no better than random chance) to 1.0 (perfect discrimination) (Hosmer et al., 2013). AUC was computed during cross-validation (10-fold with 5 repeats) on the training dataset to select optimal model parameters. Final model performance was then evaluated on the held-out test dataset. This approach allowed for fair comparison between simple models (logistic regression, LDA) and more complex techniques (SVM, boosting) while being robust to class imbalance in the dataset (Kuhn & Johnson, 2013). Confusion matrices were generated for the final selected model to report classification accuracy, sensitivity, and specificity at the 0.5 probability threshold, following the Bayes decision rule, which assigns an observation to the most likely class. This was undertaken to assess predictive performance after model selection based on cross-validated AUC.

## 3.3. Models


# 4. Results

# 4.1. Model Comparison and Selection

Eight predictive models for severe flu occurrence were evaluated: logistic regression, penalized logistic regression (elastic net), linear discriminant analysis (LDA), multivariate adaptive regression splines (MARS), generalized additive model (GAM), linear support vector machine (SVM), radial kernel SVM, and adaptive boosting (AdaBoost).Repeated 10-fold cross-validation (with 5 repeats) was implemented to objectively compare performance across the models, using AUC as the primary evaluation metric. This cross-validation approach allowed for robust comparison while minimizing the risk of overfitting. Models were then assessed on the held-out test set (20% of the original data) to validate performance and ensure the selected model would generalise appropriately to new observations from the same population.

```{r}

# Set seed for reproducibility
set.seed(299)

# Use cross validation to perform model selection
res <- resamples(list(Logistic_Regression = model.glm,
                      Penalised_Logistic_Regression = model.glmn,
                      LDA = model.lda,
                      MARS = model.mars,
                      GAM = model.gam,
                      SVM_Linear = svml.fit,
                      SVM_Radial_Kernel = svmr.fit,
                      AdaBoost = gbmA.fit))
summary(res)

bwplot(res, metric="ROC")

# Extract mean AUC values 
roc_df <- as_tibble(summary(res)$statistics$ROC, rownames = "Model")

# Select only Model and AUC
roc_clean <- roc_df %>%
  transmute(Model, AUC = round(Mean, 3))

# Display table
kable(roc_clean)

```

**Table 2. Mean cross-validated AUC (ROC) values for all candidate models. SVM with a radial kernel achieved the highest average AUC (0.701).**

The cross-validation results show that the SVM with a radial kernel achieved the highest mean AUC (0.701), with penalized logistic regression (0.696) and standard logistic regression (0.695) performing similarly.

# References

Hosmer, D. W., Lemeshow, S., & Sturdivant, R. X. (2013). Applied logistic regression (3rd ed.). John Wiley & Sons.

Kuhn, M., & Johnson, K. (2013). Applied predictive modeling. Springer.RetryClaude can make mistakes. Please double-check responses.
